Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2023-12-14 main.tex
:pencil: Update RCG method

\algrenewcommand{\algorithmicrequire}{\textbf{Input:}}
\algrenewcommand{\algorithmicensure}{\textbf{Output:}}
\algrenewcommand{\algorithmicwhile}{\textbf{While}}
\algrenewcommand{\algorithmicend}{\textbf{End}}
\algrenewcommand{\algorithmicrepeat}{\textbf{Repeat}}
\algrenewcommand{\algorithmicuntil}{\textbf{Until}}
\algrenewcommand{\algorithmicdo}{}

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}

@@

\begin{section}{Assumption}
	All proposals in this paper based on assumption of \emph{asymmetric} passive \gls{bd} \gls{ris}, i.e., symmetry constraint $\mathbf{\Theta}_g = \mathbf{\Theta}_g^\mathsf{T}$ is relaxed.
	This is feasible when asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006} are available.
	This assumption was also made in Hongyu's papers \cite{Li2023b,Li2023c}.
	For quadratic problems, the proposed algorithms may be extended to symmetric \gls{bd} \gls{ris} by replacing singular value decomposition with Takagi factorization \cite{Horn2012}.
\end{section}

\begin{section}{Point-to-Point MIMO}
	\begin{subsection}{Channel Power Maximization}
		Consider a \gls{bd} \gls{ris} with $N^\mathrm{S}$ elements, which is divided into $G$ groups of equal $L$ elements.
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\left\lVert \mathbf{H}^\mathrm{D} + \sum_g\nolimits \mathbf{H}_g^\mathrm{B} \mathbf{\Theta}_g \mathbf{H}_g^\mathrm{F} \right\rVert _\mathrm{F}^2}{\label{op:pc_power}}{}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g \in \mathcal{G} \triangleq \{1,\ldots,G\}.}{}{}
		\end{maxi!}
		For \emph{symmetric} BD-RIS, the problem has been solved in
		\begin{itemize}
			\item Matteo's paper \cite{Nerini2023}: SISO and equivalent\footnote{Single-stream MIMO with given precoder and combiner.};
			\item Ignacio's paper \cite{Santamaria2023}: SISO and directless MISO/SIMO.
		\end{itemize}

@@

		\begin{proposition}
			Start from any $\mathbf{\Theta}^{(0)}$, the sequence
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \mathbf{U}_g^{(r)} \mathbf{V}_g^{(r)}, \quad \forall g \in \mathcal{G}
			\end{equation}
			converges to a stationary point of \eqref{op:pc_power}, where $\mathbf{U}_g^{(r)}$ and $\mathbf{V}_g^{(r)}$ are left and right singular matrix of
			\begin{equation}
				\begin{split}
					\mathbf{M}_g^{(r)}
					& = {\mathbf{H}_g^\mathrm{B}}^\mathsf{H} \mathbf{H}^\mathrm{D} {\mathbf{H}_g^\mathrm{F}}^\mathsf{H} + \sum_{g' < g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r+1)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H} \\
					& \quad + \sum_{g' \ge g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H}.
				\end{split}
			\end{equation}
		\end{proposition}

@@

	\begin{subsection}{Rate Maximization}
		The problem is formulated w.r.t. precoder (instead of transmit covariance matrix) for reference:
		\begin{maxi!}
			{\scriptstyle{\mathbf{W},\mathbf{\Theta}}}{\log \det \Bigl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\sigma_n^2}\Bigr)}{\label{op:pc_rate}}{}
			\addConstraint{\lVert \mathbf{W} \rVert _\mathrm{F}^2 \le P}{}{}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g.}{}{\label{cs:pc_rate_unitary}}
		\end{maxi!}

		\eqref{op:pc_rate} is jointly non-convex and solved by \gls{ao}.
		For a given $\mathbf{\Theta}$, the optimal precoder is given by
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
		\end{equation}
		where $\mathbf{V}$ is right singular matrix of $\mathbf{H}$ and $\mathbf{S}^\star$ is a diagonal matrix of the water-filling power allocation.
		For a given $\mathbf{W}$, we update $\mathbf{\Theta}$ by \gls{rcg} method along the geodesics \cite{Abrudan2009}.
		\begin{remark}
			A geodesic refers to the shortest path between two points in a Riemannian manifold.
			% For unitary constraint \eqref{cs:pc_rate_unitary}, the
			Unitary constraint \eqref{cs:pc_rate_unitary} translates to a Stiefel manifold where the geodesics have simple expressions described by the exponential map \cite{Abrudan2008}.
			% Geodesics are the locally shortest paths in a Riemannian manifold.
			% A Geodesics are the locally shortest paths in a Riemannian manifold.
			% on Riemannian manifold are the shortest path between two points.
		\end{remark}
		For optimization problems with block unitary constraint, the adapted \gls{rcg} method at iteration $r$ for block $g$ is summarized below.
		\begin{enumerate}
			\item Compute the Euclidean gradient
			\begin{equation}
				\nabla_{\mathbf{\Theta}_g}^\mathrm{E} f^{(r)} = \frac{\partial f^{(r)}}{\partial \mathbf{\Theta}_g^*}
			\end{equation}
			\item Translate to the Riemannian gradient
			\begin{equation}
				\nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)} = \nabla_{\mathbf{\Theta}_g}^\mathrm{E} f^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathbf{\Theta}_g}^\mathrm{E} f^{(r)}}^\mathsf{H}
			\end{equation}
			\item Determine the weight factor
			\begin{equation}
				% \gamma^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl(\mathbf{G}_\mathrm{R}^{(r)} - \mathbf{G}_\mathrm{R}^{(r-1)}\bigr) \mathbf{G}_\mathrm{R}^{(r)\mathsf{H}}\Bigr)}{\mathrm{tr}\Bigl(\mathbf{G}_\mathrm{R}^{(r-1)} \mathbf{G}_\mathrm{R}^{(r-1)\mathsf{H}}\Bigr)}
				\gamma^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl(\nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)} - \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r-1)}\bigr) \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)\mathsf{H}}\Bigr)}{\mathrm{tr}\Bigl(\nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r-1)} \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r-1)\mathsf{H}}\Bigr)}
			\end{equation}
			\item Compute the conjugate direction
			\begin{equation}
				% \mathbf{D}_\mathrm{R}^{(r)} = \mathbf{G}_\mathrm{R}^{(r)} + \gamma^{(r)} \mathbf{D}_\mathrm{R}^{(r-1)}
				\mathbf{D}^{(r)} = \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)} + \gamma^{(r)} \mathbf{D}^{(r-1)}
			\end{equation}
			\item Determine the Armijo step size
			\begin{equation}
				\mu^{(r)} = \arg\max_\mu f^{(r)}\Bigl(\exp\bigl(-\mu \mathbf{D}^{(r)}\bigr) \mathbf{\Theta}_g^{(r)}\Bigr)
			\end{equation}
			\item Perform multiplicative (rotational) update
			\begin{equation}
				\mathbf{\Theta}^{(r+1)} = {\color{gray}\exp}\Bigl(-\mu^{(r)} \mathbf{D}^{(r)}\Bigr) \mathbf{\Theta}_g^{(r)}
			\end{equation}
		\end{enumerate}
		The advantage over general manifold optimization \cite{Absil2009,Pan2022d} is trifold: 1) no retraction is required; 2) no projection is required; 3) no matrix inversion is required.
	\end{subsection}
\end{section}

>>>>

\algrenewcommand{\algorithmicrequire}{\textbf{Input:}}
\algrenewcommand{\algorithmicensure}{\textbf{Output:}}
\algrenewcommand{\algorithmicwhile}{\textbf{While}}
\algrenewcommand{\algorithmicend}{\textbf{End}}
\algrenewcommand{\algorithmicrepeat}{\textbf{Repeat}}
\algrenewcommand{\algorithmicuntil}{\textbf{Until}}
\algrenewcommand{\algorithmicfor}{\textbf{For}}
\algrenewcommand{\algorithmicdo}{}

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}

@@

\begin{section}{Assumption}
	We introduce \gls{bd} \gls{ris} in \gls{mimo} \gls{pc} and \gls{ic}.
	% point-to-point and interference \gls{mimo} channels.
	All proposals are based on assumption of \emph{asymmetric} passive \gls{bd} \gls{ris}, i.e., symmetry constraint $\mathbf{\Theta}_g = \mathbf{\Theta}_g^\mathsf{T}$ is relaxed.
	This is feasible when asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006} are available.
	This assumption was also made in Hongyu's papers \cite{Li2023b,Li2023c}.
	For quadratic problems, the proposed algorithms may be extended to symmetric \gls{bd} \gls{ris} by replacing singular value decomposition with Takagi factorization \cite{Horn2012}.
\end{section}

\begin{section}{\glsentryshort{mimo}-\glsentryshort{pc}}
	\begin{subsection}{Channel Power Maximization}
		Consider a \gls{bd} \gls{ris} with $N^\mathrm{S}$ elements, which is divided into $G$ groups of equal $L$ elements.
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\left\lVert \mathbf{H}^\mathrm{D} + \sum_g\nolimits \mathbf{H}_g^\mathrm{B} \mathbf{\Theta}_g \mathbf{H}_g^\mathrm{F} \right\rVert _\mathrm{F}^2}{\label{op:pc_power}}{}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g \in \mathcal{G} \triangleq \{1,\ldots,G\}.}{}{}
		\end{maxi!}
		For \emph{symmetric} BD-RIS, the problem has been solved in
		\begin{itemize}
			\item Matteo's paper \cite{Nerini2023}: SISO and equivalent\footnote{Single-stream MIMO with given precoder and combiner.};
			\item Ignacio's paper \cite{Santamaria2023}: SISO and directless MISO/SIMO.
		\end{itemize}

@@

		\begin{proposition}
			Start from any $\mathbf{\Theta}^{(0)}$, the sequence
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \mathbf{U}_g^{(r)} \mathbf{V}_g^{(r)}, \quad \forall g
			\end{equation}
			converges to a stationary point of \eqref{op:pc_power}, where $\mathbf{U}_g^{(r)}$ and $\mathbf{V}_g^{(r)}$ are left and right singular matrix of
			\begin{equation}
				\begin{split}
					\mathbf{M}_g^{(r)}
					& = {\mathbf{H}_g^\mathrm{B}}^\mathsf{H} \mathbf{H}^\mathrm{D} {\mathbf{H}_g^\mathrm{F}}^\mathsf{H} + \sum_{g' < g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r+1)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H} \\
					& \quad + \sum_{g' \ge g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H}.
				\end{split}
			\end{equation}
		\end{proposition}

@@

	\begin{subsection}{Rate Maximization}
		The problem is formulated w.r.t. precoder (instead of transmit covariance matrix) for reference:
		\begin{maxi!}
			{\scriptstyle{\mathbf{W},\mathbf{\Theta}}}{\log \det \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\sigma_n^2}\biggr)}{\label{op:pc_rate}}{\label{ob:pc_rate}}
			\addConstraint{\lVert \mathbf{W} \rVert _\mathrm{F}^2 \le P}{}{}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g.}{}{\label{cs:pc_rate_unitary}}
		\end{maxi!}

		\eqref{op:pc_rate} is jointly non-convex and solved by \gls{ao}.
		For a given $\mathbf{\Theta}$, the optimal precoder is given by
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
		\end{equation}
		where $\mathbf{V}$ is right singular matrix of $\mathbf{H}$ and $\mathbf{S}^\star$ is a diagonal matrix of the water-filling power allocation.
		For a given $\mathbf{W}$, we update $\mathbf{\Theta}$ by \gls{rcg} method along the geodesics \cite{Abrudan2009}.
		\begin{remark}
			A geodesic refers to the shortest path between two points in a Riemannian manifold.
			% For unitary constraint \eqref{cs:pc_rate_unitary}, the
			Unitary constraint \eqref{cs:pc_rate_unitary} translates to a Stiefel manifold where the geodesics have simple expressions described by the exponential map \cite{Abrudan2008}.
			% Geodesics are the locally shortest paths in a Riemannian manifold.
			% A Geodesics are the locally shortest paths in a Riemannian manifold.
			% on Riemannian manifold are the shortest path between two points.
		\end{remark}
		For general optimization problems with block unitary constraint, the adapted \gls{rcg} method at iteration $r$ for block $g$ is summarized below, where $f^{(r)}$ is evaluated over $\bigl\{ \{\mathbf{\Theta}_{g'}^{(r+1)}\}_{g'=1}^{g-1}, \{\mathbf{\Theta}_{g'}^{(r)}\}_{g'=g}^{G} \bigr\}$.
		\begin{enumerate}
			\item Compute the Euclidean gradient
			\begin{equation}
				\nabla_g^\mathrm{E} f^{(r)} = \frac{\partial f^{(r)}}{\partial \mathbf{\Theta}_g^*};
				\label{eq:rcg_gradient_euclidean}
			\end{equation}
			\item Translate to the Riemannian gradient
			\begin{equation}
				\nabla_g^\mathrm{R} f^{(r)} = \nabla_g^\mathrm{E} f^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {\nabla_g^\mathrm{E} f^{(r)}}^\mathsf{H};
				\label{eq:rcg_gradient_riemannian}
			\end{equation}
			\item Determine the weight factor
			\begin{equation}
				\gamma_g^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl(\nabla_g^\mathrm{R} f^{(r)} - \nabla_g^\mathrm{R} f^{(r-1)}\bigr) \nabla_g^\mathrm{R} {f^{(r)}}^\mathsf{H}\Bigr)}{\mathrm{tr}\Bigl(\nabla_g^\mathrm{R} f^{(r-1)} \nabla_g^\mathrm{R} {f^{(r-1)}}^\mathsf{H}\Bigr)};
				\label{eq:rcg_weight}
			\end{equation}
			\item Compute the conjugate direction
			\begin{equation}
				\mathbf{D}_g^{(r)} = \nabla_g^\mathrm{R} f^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)};
				\label{eq:rcg_direction}
			\end{equation}
			\item Determine the Armijo step size\footnote{To double the step size, simply square the argument instead of recomputing the matrix exponential, i.e., $\exp(2 \mu_g \mathbf{D}_g) = \exp^2(\mu_g \mathbf{D}_g)$.}
			\begin{equation}
				\mu_g^{(r)} = \arg\max_{\mu_g} f^{(r)}\Bigl(\exp\bigl(\mu_g \mathbf{D}_g^{(r)}\bigr) \mathbf{\Theta}_g^{(r)}\Bigr);
				\label{eq:rcg_step}
			\end{equation}
			\item Perform rotational update along local geodesics
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \exp\Bigl(\mu_g^{(r)} \mathbf{D}_g^{(r)}\Bigr) \mathbf{\Theta}_g^{(r)}.
				\label{eq:rcg_update}
			\end{equation}
		\end{enumerate}

		\begin{remark}
			The adapted \gls{rcg} method leverages the fact that block unitary matrices are closed under multiplication (but not necessarily under addition).
			Its advantage over universal manifold optimization \cite{Absil2009,Pan2022d} is trifold:
			\begin{itemize}
				\item No retraction is involved;
				\item Lower computational complexity per iteration \cite{Abrudan2008};
				\item Faster convergence thanks appropriate operational space.
			\end{itemize}
		\end{remark}

		The complex derivative of \eqref{ob:pc_rate} w.r.t. \gls{ris} block $g$ is
		\begin{equation}
			\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\sigma_n^2} {\mathbf{H}_g^\mathrm{B}}^\mathsf{H} \mathbf{HW} \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\sigma_n^2}\biggr)^{-1} \mathbf{W}^\mathsf{H} {\mathbf{H}_g^\mathrm{F}}^\mathsf{H}.
			\label{eq:pc_rate_gradient_ris}
		\end{equation}
		The adapted \gls{rcg} method for the \gls{ris} rate maximization subproblem is summarized in Algorithm~\ref{ag:pc_rate_ris}.

		\setalgorithmcaptionfont{\small}
		\begin{algorithm}[!t]
			\small
			\caption{\gls{rcg} Method for \gls{ris} \gls{mimo}-\gls{pc} Rate Maximization}
			\label{ag:pc_rate_ris}
			\begin{algorithmic}[1]
				\Require $\mathbf{H}^\mathrm{D}$, $\mathbf{H}^\mathrm{F}$, $\mathbf{H}^\mathrm{B}$, $\mathbf{W}$, $L$, $\sigma_n^2$
				\Ensure $\mathbf{\Theta}^\star$
				% \State Initialize $r \gets 0$, $\mathbf{\Theta}^{(0)}$
				\State $r \gets 0$, $\mathbf{\Theta}^{(0)}$
				\Repeat
					% \State Update $r \gets r+1$
					\State $r \gets r+1$
					\For {$g \gets 1$ to $G$}
						% \State Get $\mathbf{\Theta}_g^{(r)}$ by \eqref{eq:pc_rate_gradient_ris}, \eqref{eq:rcg_gradient_riemannian} -- \eqref{eq:rcg_update}
						\State $\mathbf{\Theta}_g^{(r)} \gets$ \eqref{eq:pc_rate_gradient_ris}, \eqref{eq:rcg_gradient_riemannian}--\eqref{eq:rcg_update}
					\EndFor
				\Until $\lvert R^{(r)} - R^{(r-1)} \rvert / R^{(r-1)} \le \epsilon$
			\end{algorithmic}
		\end{algorithm}
	\end{subsection}
\end{section}
